### 2012-09-18 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updates to linear problem chapter

The purpose of this chapter is to provide a full background on solving
linear systems as it relates to this work. It may seem like there is a
lot here, but we really do need it to fully define our new methods and
compare both qualitatively and quantitatively to convential
methods. For example, we need a solid explanation of stationary
methods because MCSA is in fact an acceleration of a stationary method
and therefore shares many properties with them. Subspace methods are
the current class of methods most widely used for sparse systems and
are at the core of the Newton-Krylov methods that we will be comparing
the FANM method with. The most rigorous piece of this chapter should
of course be the stochastic solver definitions, also providing me a
place to include some of my work in this area. Finally,
preconditioning is important for this work, MCSA requires it and
Newton-Krylov methods almost always need some type of
preconditioning. Therefore, we must also discuss these aspects.

@@

Choosing GMRES as our model Krylov method, we are first tasked with
finding a projector onto the subspace. We seek an orthonormal basis
for $\mathcal{K}_m(\ve{A},\ve{r}_0)$ by an orthogonalization procedure
that is commonly based on, but not limited to, the \textit{Arnoldi}
recurrence relation. The Arnoldi procedure will generate an
orthonormal basis, $\ve{V}_m \in \mathbb{C}^{N \times m}$, via a
variant of the Gram-Schmidt procedure that re-applies the operator for
each consecutive vector, thus forming a basis that spans the subspace
in Eq~(\ref{eq:krylov_subspace}). Due to its equivalent
dimensionality, $m$, to that of the subspace, we will refer to such
recurrence relations as \textit{long recurrence relations}. Those
orthogonal projection procedures that have a dimensionality less than
$m$ will be refferred to as \textit{short recurrence relations}.  Once
$\ve{V}_m$ is found, per the constraint subspace definition it then
follows that its basis is defined as $\ve{W}_m = \ve{A} \ve{V}_m$.

Knowing the projections onto the search and constraint subspaces, the
GMRES method may be formulated. We note here several properties of
this formulation and how they may facilitate or hinder the solution of
large-scale, sparse linear problems, also noting that these properties
are common among many Krylov methods. First, from a memory perspective
GMRES is efficient in that the operator $\ve{A}$ need not be
explicitly stored. Rather, only the ability to compute the action of
that operator on a vector of valid size is required. However, these
savings in memory are balanced by the fact that the long recurrence
relations used in the Arnoldi procedure require all vectors that span
the Krylov space to be stored. If the size of these vectors becomes
prohibitive, the Arnoldi procedure can be restarted at the cost of
losing information in the orthogonalization process, creating the
potentitial to generate new search directions that are not orthogonal
to all previous search directions (and therefore less than
optimal). From an implementation perspective, because the operator is
not required to be formed, GMRES is significantly more flexible in its
usage in that there are many instances where various processes serve
to provide the action of that operator (e.g. radiation transport
sweeps \citep{evans_2010}) that may not be amenable to its
construction.

@@

\subsubsection{Adjoint Method}
\label{subsubsec:adjoint_mc}
An alternative formulation for Monte Carlo matrix inversion is the
adjoint method. In this scheme, random walk histories are born in a
state generated by source sampling and tallies are generated in the
states in which the random walk currently resides. In terms of
particle transport, this adjoint method is equivalent to a traditional
forward method. To begin, we first form the \textit{adjoint
  Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^T = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where now we are forming the decomposition with respect to the
transpose of $\ve{H}$. We then following the same procedure as the
direct method for forming the probability and weight matrices in the
decomposition. Using the transpose form, probabilities should instead
by column-scaled:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state $j$ from the current state
in the random walk $j$ by sampling column-wise. Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
the weight generated for a particular random walk permutation as in
Eq~(\ref{eq:mc_walk_permutation}) and the fact that in the adjoint
method we are only tallying the solution in the state in which the
walk currently resides, the contribution to the solution from a
particular random walk permutation is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} b_{i_0} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $b_{i_0}$ is that sampled source starting
state. Note here that the estimator in
Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
dependency on the source state as in
Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
the situation in the direct method where we much start a random walk
in each source state for every permutation such that we may compute a
contribution for that state. In the adjoint method, we instead tally
in all states and those of lesser importance will not be visited as
frequently by the random walk. Finally, the estimator using all
permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded in the same way as the direct method, directly
recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k} \\ &= x_i\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimator.

@@

@@

\section{Parallel Methods for Solving Sparse Linear Systems}
\label{sec:parallel_krylov_methods}
Modern parallel implementations of Krylov methods on distributed
memory architectures rely heavily on capabilities provided by general
linear algebra frameworks. This arises from the fact that Krylov
methods require only a handful of operations in their
implementation. Per Saad's text \citep{saad_2003}, these operations
are preconditioning, matrix-vector multiplications, vector updates,
and dot products. For the last three items, linear algebra libraries
such as PETSc \citep{petsc_2012} and Trilinos \citep{trilinos_2005}
provide efficient parallel implementations for these
operations. Depending on the type of preconditioning used, efficient
parallel implementations may also be available for those
operations. In all cases, a series of scatter/gather operations are
required such that global communication operations must
occur. Therefore, the performance of such operations is bound to the
implementation. However, asymptotically perfomance should be the same
across all implementations.

>>>>

The purpose of this chapter is to provide a full background on solving
linear systems as it relates to this work. It may seem like there is a
lot here, but we really do need it to fully define our new methods and
compare both qualitatively and quantitatively to convential
methods. For example, we need a solid explanation of stationary
methods because MCSA is in fact an acceleration of a stationary method
and therefore shares many properties with them. Subspace methods are
the current class of methods most widely used for sparse systems and
are at the core of the Newton-Krylov methods that we will be comparing
the FANM method with. The most rigorous piece of this chapter should
of course be the stochastic solver definitions, also providing me a
place to include some of my work in this area. Finally,
preconditioning is important for this work, MCSA requires it and
Newton-Krylov methods almost always need some type of
preconditioning. Therefore, we must also discuss these
aspects.\footnote{The contents of this chapter, particularly those
  sections relating to projection methods, are heavily based on Saad's
  text \citep{saad_2003}.}

@@

Choosing GMRES as our model Krylov method, we are first tasked with
finding a projector onto the subspace. We seek an orthonormal basis
for $\mathcal{K}_m(\ve{A},\ve{r}_0)$ by an orthogonalization procedure
that is commonly based on, but not limited to, the \textit{Arnoldi}
recurrence relation. The Arnoldi procedure will generate an
orthonormal basis, $\ve{V}_m \in \mathbb{C}^{N \times m}$, via a
variant of the Gram-Schmidt procedure that re-applies the operator for
each consecutive vector, thus forming a basis that spans the subspace
in Eq~(\ref{eq:krylov_subspace}). Due to its equivalent
dimensionality, $m$, to that of the subspace, we will refer to such
recurrence relations as \textit{long recurrence relations}. Those
orthogonal projection procedures that have a dimensionality less than
$m$ will be refferred to as \textit{short recurrence relations}.  Once
$\ve{V}_m$ is found, per the constraint subspace definition it then
follows that its basis is defined as $\ve{W}_m = \ve{A}
\ve{V}_m$. Knowing the projections onto the search and constraint
subspaces, the GMRES iteration may be formulated as follows:
\begin{algorithm}[htpb!]
  \caption{GMRES Iteration}
  \label{alg:gmres}
  \begin{algorithmic}
    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
    \State $\beta := ||\ve{r}_0||_2$
    \State $\ve{v}_1 := \ve{r}_0 / \beta$
    \Comment{Create the orthonormal basis for the Krylov subspace}
    \For{$j = 1, 2, \cdots, m$}
    \State $h_{ij} \leftarrow \langle w_j,v_j \rangle$
    \State $w_j \leftarrow w_j - h_{ij}v_i$
    \EndFor
    \State $h_{j+1,j} \leftarrow ||w_j||_2$
    \State $v_{j+1} \leftarrow w_j / h_{j+1,j}$
    \Comment{Apply the orthogonality constraints}
    \State $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 - \ve{H}_m\
    \ve{y}||_2 $
    \State $\ve{x}_m \leftarrow \ve{x}_0 + \ve{V_m} \ve{y}_m$
  \end{algorithmic}
\end{algorithm}

We note here several properties of this formulation and how they may
facilitate or hinder the solution of large-scale, sparse linear
problems, also noting that these properties are common among many
Krylov methods. First, from a memory perspective GMRES is efficient in
that the operator $\ve{A}$ need not be explicitly stored. Rather, only
the ability to compute the action of that operator on a vector of
valid size is required. However, these savings in memory are balanced
by the fact that the long recurrence relations used in the Arnoldi
procedure require all vectors that span the Krylov space to be
stored. If the size of these vectors becomes prohibitive, the Arnoldi
procedure can be restarted at the cost of losing information in the
orthogonalization process, creating the potentitial to generate new
search directions that are not orthogonal to all previous search
directions (and therefore less than optimal). From an implementation
perspective, because the operator is not required to be formed, GMRES
is significantly more flexible in its usage in that there are many
instances where various processes serve to provide the action of that
operator (e.g. radiation transport sweeps \citep{evans_2010}) that may
not be amenable to its construction.

@@

\subsubsection{Adjoint Method}
\label{subsubsec:adjoint_mc}
An alternative formulation for Monte Carlo matrix inversion is the
adjoint method. In this scheme, random walk histories are born in a
state generated by source sampling and tallies are generated in the
states in which the random walk currently resides. In terms of
particle transport, this adjoint method is equivalent to a traditional
forward method. To begin, we first form the \textit{adjoint
  Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^T = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where now we are forming the decomposition with respect to the
transpose of $\ve{H}$. We then following the same procedure as the
direct method for forming the probability and weight matrices in the
decomposition. Using the transpose form, probabilities should instead
by column-scaled:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state $j$ from the current state
in the random walk $j$ by sampling column-wise. Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
the weight generated for a particular random walk permutation as in
Eq~(\ref{eq:mc_walk_permutation}) and the fact that in the adjoint
method we are only tallying the solution in the state in which the
walk currently resides, the contribution to the solution from a
particular random walk permutation is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} b_{i_0} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $b_{i_0}$ is that sampled source starting
state. Note here that the estimator in
Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
dependency on the source state as in
Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
the situation in the direct method where we must start a random walk
in each source state for every permutation such that we may compute a
contribution for that state. In the adjoint method, we instead tally
in all states and those of lesser importance will not be visited as
frequently by the random walk. Finally, the estimator using all
permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded in the same way as the direct method, directly
recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k} \\ &= x_i\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimator.

@@

Given these ideas, we can then generate an algorithm for implementing
the adjoint method:
\begin{algorithm}[htpb!]
  \caption{Adjoint Monte Carlo}
  \label{alg:adjoint_mc}
  \begin{algorithmic}
    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
  \end{algorithmic}
\end{algorithm}

@@

\section{Parallel Methods for Solving Sparse Linear Systems}
\label{sec:parallel_krylov_methods}
Modern parallel implementations of methods on distributed memory
architectures rely heavily on capabilities provided by general linear
algebra frameworks. For methods like GMRES, this arises from the fact
that Krylov methods require only a handful of operation types in their
implementation that can be efficiently programmed on these
architectures. Per Saad's text \citep{saad_2003} and as noted in
Algorithm~\ref{alg:gmres}, these operations are preconditioning,
matrix-vector multiplications, vector updates, and inner products. For
the last three items, linear algebra libraries such as PETSc
\citep{petsc_2012} and Trilinos \citep{trilinos_2005} provide
efficient parallel implementations for these operations. Depending on
the type of preconditioning used, efficient parallel implementations
may also be available for those operations. In all cases, a series of
scatter/gather operations are required such that global communication
operations must occur. Although the relative performance of such
operations is bound to the implementation, asymptotically perfomance
should be the same across all implementations.

### 2012-09-18 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
updates to linear problem chapter

This work will result in a new and novel nonlinear solver, FANM. To
movitate the development of FANM and thoroughly understand both its
mathematical properties and its qualities as related to other
nonlinear methods, we need a full workup of current iterative methods
for solving nonlinear systems. We should acknowledge operator split
and linearization methods for dealing with nonlinear systems
(e.g. think pressure correction strategies for Navier-Stokes) but the
focus here should be on fully-implicit nonlinear solves that are
amenable to solving tightly-coupled multi-physics systems in a way that
is fully consistent.

The key here will be a full understanding of Newton methods (and they
actually are quite simple compared to some of the linear solver
methods like subspace methods). It is important to understand how the
correction is computed from the Jacobian and how we may improve this
with globalization methods. Although I may not use globalization
methods in my work, they are important enough in practice that we must
be sure that a FANM method is not prohibited from using these. Current
methods, primarily Newton-Krylov methods, will of primary concern here
as these will serve as the baseline for assessing the qualities of a
FANM solver.

Finally, as we will choose transient problems for our models, time
integration in the context of nonlinear systems will be consider and
we will survey applications to model problems in the field ( there is
a vast amount of literature on fully-implicit nonlinear solves for the
Navier-Stokes equations) that will give good benchmarks to test (such
as natural convection and driven cavity problems).

>>>>

This work will result in a new and novel nonlinear solver, FANM. To
movitate the development of FANM and thoroughly understand both its
mathematical properties and its qualities as related to other
nonlinear methods, we need a full workup of current iterative methods
for solving nonlinear systems. We should acknowledge operator split
and linearization methods for dealing with nonlinear systems
(e.g. think pressure correction strategies for Navier-Stokes) but the
focus here should be on fully-implicit nonlinear solves that are
amenable to solving tightly-coupled multiphysics systems in a way that
is fully consistent.

The key here will be a full understanding of Newton methods (and they
actually are quite simple compared to some of the linear solver
methods like subspace methods). It is important to understand how the
correction is computed from the Jacobian and how we may improve this
with globalization methods. Although I may not use globalization
methods in my work, they are important enough in practice that we must
be sure that a FANM method is not prohibited from using these. Current
methods, primarily Newton-Krylov methods, will of primary concern here
as these will serve as the baseline for assessing the qualities of a
FANM solver. Finally, we will survey applications to model problems in
the field ( there is a vast amount of literature on fully-implicit
nonlinear solves for the Navier-Stokes equations) that will give good
benchmarks to test (such as natural convection and driven cavity
problems).
